Fluent statistical computing interfaces for biological data analysis

### 2020-02-23 Rmd/05-embeddings.Rmd
add outline of liminal paper

# Spurious embeddings from non-linear dimensionality reduction methods and how to find them {#ch:tsne}

>>>>

# Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings {#ch:tsne}

<!--
- review of embedding algorithms, interactive graphics, tours
- technology
- design
- examples/benchmarks
  - Kevin moon
  - pdfsense
  - single cell
  - quick draw
https://jlmelville.github.io/smallvis/init.html#other_initialization_options
-->

## Introduction

High dimensional data is increasingly prevalent in the natural sciences and
beyond but presents a challenge to the analyst in terms of both data cleaning /
preprocessing and visualisation. Methods to embed data from a high-dimesional
space into a low-dimensional one now form a core step of the data analysis
workflow where they are used to ascertain hidden structure and denoise data for
downstream analysis (thereby nullifying the 'curse of dimensionality').

Choosing an appropriate embedding presents a challenge to the analyst. How does
an analyst know whether the embedding has captured the underlying topology and
geometry of the high dimensional space? The answer depends on the analyst's
workflow.  @Brehmer2014-hk characterised two main workflow steps that an
analyst performs when using embedding techniques: dimension reduction and
cluster orientation. The first relates to dimension reduction achieved by using
an embedding method, here an analyst wants to characterise and map meaning onto
the embedded form, for example identifying batch effects from a high throughput
sequencing experiment, or identifying a gradient or trajectory along the
embedded form @Nguyen2019-yh. The second relates to using embeddings as part of
a clustering workflow. Here analysts are interested in identifying and naming
clusters and verifying them by either applying known labels or colouring by
variables that are a-priori known to distinguish clusters. Both of these
workflow steps rely on the embedding being 'faithful' or the original high
dimensional dataset, and become much more difficult when there is no underlying
ground truth.

Embedding methods can be classified into two broad groups: linear and
non-linear methods. Linear methods perform a linear transformation of the data;
one example is principal components analysis (PCA) which performs an
eigendecomposition of the estimated sample covariance matrix. The eigenvalues
are sorted in decreasing order and represent the variance explained by each
component (eigenvector). A common approach to deciding on the number of
principal components to retain is to plot the proportion of variance explained
by each component and choose a cut-off.

Non-linear methods generally perform preprocessing on the high-dimensional data
such as generating a neighborhood graph and perform transformations on the
preprocessed form. We restrict our attention to three methods that are commonly
used in high-throughput biology: t-distributed stochastic neighbor embedding
(t-SNE), uniform manifold alignment and projection (UMAP), and potential of
heat-diffusion for affinity-based transition embedding (PHATE). The t-SNE
algorithm estimates the similarity of (Euclidean) distances of points in a high
dimensional space using a Gaussian distribution and then estimates a
configuration in the low dimensional embedding space by modelling similarities
using a t-distribution with 1 degree of freedom. The resulting configuration is
the one that minimizes the Kulback-Leibler divergence between the two
distributions. A recent theoretical contribution by @Linderman2019-dq proved
that t-SNE can recover spherical and well separated cluster shapes, and
proposed new approaches for tuning the optimisation parameters. It is a known
problem that t-SNE can have trouble recovering global structure and that
configurations can be highly dependent on how the algorithm is initialised and
parameterised [@Wattenberg2016-ji; @Kobak2019-lm]. UMAP is a method that is
related to LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest
neighbor graph. Its main differences are that it uses a different cost function
(cross entropy) which is optimized using stochastic gradient descent and
defines a different kernel for similiarites in the low dimensional space. Due
to it's computational speed it's possible to generate UMAP embeddings in more
than three dimensions. Finally, PHATE, inspired by diffusion maps, is based on
estimating an affinity matrix via a distance matrix and k-nearest neighbors
graph. The algorithm denoises estimated distances in high dimensional space via
transforming the affinity matrix into a Markov transition probaility matrix and
diffusing this matrix over a fixed number of time steps. Then the diffused
probabilities are transformed once more to construct a distance matrix, and
multidimensional scaling is performed to construct a 2d embedding for
visualization.

As part of a visualization workflow, it's important to consider the perception
and interpretation of embedding methods as well. @Sedlmair2013-pn showed that
2D scatter plots were mostly suffucient for detecting class seperation, however
they noted that often multiple embeddings were required. For the task of
cluster identification, @Lewis2012-ai showed experimentally that novice users
of non-linear embedding techniques were more likely to consider clusters of
points on a 2d scatter plot to be the result of a spurious embedding compared
to advanced users who were aware of the inner workings of the embedding
algorithm.

A complimentary approach for visualizing structure in high dimensional data is
the tour. A tour is a sequence of projections of a high dimensional dataset
onto a low-dimensional orthonormal basis matrix, that is represented as a
dynamic visualization. The sequence of generated bases are interpolated to form
the tour path, allowing a user to explore the subspace of projections. A grand
tour corresponds to choosing new bases at random, and can give an overview of
the structure in the data. Instead of picking projections at random, a guided
tour can be used to generate a sequence 'interesting' projections as quantified
by an index function. Given the dynamic nature of the tour, user interaction is
important for controlling and exploring the visualisation: the tour has been
used previously by @Wickham2015-cx as tool for exploring statistical model fits
and by @Buja1996-fk for exploring factorial experimental designs.

While there has been much work on the algorithmic details of the aformentioned
embedding methods, there has been relatively few tools designed to assist users
to interact with these techniques and assist them in making comparsions between
embeddings and performing the aforementioned cluster orientation tasks. Several
interactive interfaces have been proposed for evaluating or using embedding
techniques: the Sleepwalk interface provides a click and highlight
visualisation for colouring points in an embedding according to their distance
in the original high-dimensional space [@Ovchinnikova2019-gf]. The work by
@Pezzotti2017-cz provides a user guided and modified form of the t-SNE
algorithm, that allows users to modified optimisation parameters in real-time.
Similarly, the embedding projector is a web interface to running UMAP, t-SNE or
PCA live in the browser and provides interactions to color points, and
highlights nearest neighbors [@Smilkov2016-hp].

There is no one-size fits all: finding an appropriate embedding for a given
dataset is a difficult and somewhat poorly defined problem. For non-linear
methods, there are a lot of parameters to explore that can have an effeft on
the resulting visualisation and interpretation. Interfaces for evaluating
embeddings require interaction but should also be able to be incorporated into
an analysts workflow. We propose a more pragmatic workflow inspired by
incorporating interactive graphics and tours with embeddings that allows users
to see a global overview of their high dimensional data and assists them with
cluster orientation tasks. This workflow is incorporated into an R package
called `liminal` (Available: https://github.com/sa-lee/liminal).

## Design

We propose using tours as part of an analyst's workflow in in performing
dimensionality reduction tasks. We have made extensive use of ensemble
graphics, that is aligning related plots alongside each other to provide
context. As we will see in the case studies, this allows analysts to quickly
compare views from embedding methods and allows them to see how the embedding
method alters the global structure of their data. Using ensembles allows the
use of interaction techniques, that allow analysts to perform cluster
orientation tasks via linking multiple views. This approach allows our
interface, to achieve the three principles for interactive high-dimensional
data visualisation outlined by @Buja1996-fk: finding gestalt, posing queries,
and making comparisons.

### Finding Gestalt: focus and context

To investigate latent structure and the shape of a high dimensional dataset, a
tour can be run without the use of an external embedding. It is often useful to
first run principal components on the input as an intial dimension reduction
step, and then tour a subset of those components instead, i.e. by selecting
them from a scree plot. The default tour layout is a scatter plot with an axis
layout displaying the magnitude and direction of each basis vector. Since the
tour is dynamic, it is often useful to be able to pause and highlight a
particular view. In our interface, brushing will pause the tour path, allowing
users to identify 'interesting' projections. The domain of the axis scales from
running a tour is called the half range, and is computed by rescaling the input
data onto hyperdimensional unit cube. We bind the half range to a mouse wheel
event, allowing a user to pan and zoom on the tour view dynamically. This is
useful for peeling back dense clumps of points to reveal structure.

### Posing Queries: multiple views, many contexts

We combine the tour view in a side by side layout with a scatter
plot view from an embedding algorithm. The views are linked; analysts can brush a group of points in the embedding view, and see its shape in the tour view.
This can be used for either manually identifying clusers, or verifying the location of clusters relative to other groupings. Consequently, simple linked
brushing can alleviate a downfall of non-linear methods such as UMAP or t-SNE
that break hierachial structure of clusterings. Aside from simple
linked brushing, we can also pose more complex queries via interrogating
the neighbourhood graph. Using a linked neighbourhood brush, we can 
visually investigate the nearest neighbour relationships in the 
high-dimensional space via brushing in the embedding view. The user can select
the number of nearest neighbours directly, and modify the distance metric 
used for determining the neighbours. This allows users to interrogate the stability of clusters generated in the embedding view. Multiple brushes
can be used to pose queries in either the tour view or the embedding view;
interface controls allow these brushes to combine using logical operators
such as 'and', 'or', or 'not'. The use of linked brushing goes beyond
single point highlights or colorings, allowing analysts to get a more holistic
view of the effect of an embedding algorithm.

### Making comparsions

TODO

## Implementation

The `liminal` software is implemented as an open source R package @baseR. All
visualisations produced by the package were constructed using the `vegawidget`
package, which is an R interface to the Vega-Lite grammar of interactive
graphics. Ensemble graphics for arranging tour and embedding views are
generated using the `shiny` packages. We have made extensive use of the `tourr`
package for constructing tour paths.

### Tours as a streaming data problem

### Linking views via spatial brushes

## Case Studies

### Case Study 1: Fake trees

### Case Study 2: well-clustered smallish single cell

### Case Study 3: largeish single cell data

### Case Study 4: non-spherical clusters, pdfsense

## Discussion
